---
title: "CamCAN Selectivity & Attrition Analysis (for *The Cambridge Centre for Ageing and Neuroscience (Cam-CAN) longitudinal study protocol: Phase 4 (Enrichment) and Phase 5 (Rescan)*)"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2026-05-15"
author: Ina Demetriou & Richard Henson 
---

```{r}
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(grid)
library(cowplot)
library(GGally)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(isotree)
library(lmerTest)
library(irr)
```

#Plotting function helper 
```{r}
plot_normality <- function(data, column_name) {
  column_data <- data[[column_name]]
  data_clean <- na.omit(column_data)
    density_estimate <- density(data_clean)
    normal_max <- dnorm(mean(data_clean), mean = mean(data_clean), sd = sd(data_clean))
    ylim_max <- max(density_estimate$y, normal_max)
    hist(data_clean,  main = NA, xlab = column_name, ylab = "Density", 
         breaks = 20, col = "skyblue", border = "white", probability = TRUE,
         ylim = c(0, ylim_max * 1.1),
         las = 1) 
    lines(density_estimate, col = "blue", lwd = 2, lty = 1)
    curve(dnorm(x, mean = mean(data_clean), sd = sd(data_clean)), 
          col = "red", lwd = 2, lty = 2, add = TRUE)
    legend("topright", inset = c(0, -0.2), legend = c("Density", "Normal Distribution"), col = c("blue", "red"), lwd = 2, bty = "n", bg = "white", xpd = TRUE)
}

plot_violin <- function(data, column_name, group_column) {
  # Remove missing values and prepare the data
  data_clean <- data[!is.na(data[[column_name]]) & !is.na(data[[group_column]]), ]
  data_clean[[column_name]] <- scale(data_clean[[column_name]])
  # Create violin plot with grouping
  ggplot(data_clean, aes_string(x = group_column, y = column_name, fill = group_column)) +
    geom_violin(trim = FALSE, alpha = 0.7) +
    geom_boxplot(width = 0.1, position = position_dodge(0.9), alpha = 0.5 ) +
    scale_fill_manual(
      values = c("0" = "lightblue", "1" = "darkorange")  # Custom colors
    ) +
    ylim(-4, 4) +
    scale_x_discrete(labels = c("0" = "NR", "1" = "R")) +  # Relabel x-axis
    labs(x = NULL, y = column_name) +  # Remove legend title
    theme_minimal() +
    theme(legend.position = "none")  # Remove the legend
}

plot_violin_outlier_suppress <- function(data, column_name, group_column) {
  # Remove missing values and prepare the data
  data_clean <- data[!is.na(data[[column_name]]) & !is.na(data[[group_column]]), ]
  data_clean[[column_name]] <- scale(data_clean[[column_name]])
  # Create violin plot with grouping
  ggplot(data_clean, aes_string(x = group_column, y = column_name, fill = group_column)) +
    geom_violin(trim = FALSE, alpha = 0.7) +
    geom_boxplot(width = 0.1, position = position_dodge(0.9), alpha = 0.5, outlier.shape = NA ) +
    scale_fill_manual(
      values = c("0" = "lightblue", "1" = "darkorange")  # Custom colors
    ) +
    ylim(-3.5, 3.5) +
    scale_x_discrete(labels = c("0" = "NR", "1" = "R")) +  # Relabel x-axis
    labs(x = NULL, y = column_name) +  # Remove legend title
    theme_minimal() +
    theme(legend.position = "none")  # Remove the legend
}
plot_outliers <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, outliers) {
  outlier_CCIDs <- unique(outliers$CC.ID)
  data <- data %>%
    mutate(is_outlier = ifelse(CC.ID %in% outlier_CCIDs, TRUE, FALSE))
  
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(aes(color = is_outlier, alpha = is_outlier), size = 2, stroke = 1) +
    geom_line(aes(group = CC.ID), lty = 2, alpha = 0.2, color = "black") +
    geom_smooth(
      aes(group = CC.ID, color = is_outlier),
      method = "lm", se = FALSE, size = 0.5, lty = 1
    ) +
    labs(x = x_label, y = y_label) +
    ylim(ymin, ymax) +
    scale_color_manual(
      values = c("TRUE" = "green", "FALSE" = "darkorange"),
      labels = c("TRUE" = "Outlier", "FALSE" = "Non-Outlier")
    ) +
    scale_alpha_manual(values = c("TRUE" = 0.3, "FALSE" = 0.1), guide = "none") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "bottom",
      legend.box = "horizontal"
    ) +
    guides(color = guide_legend(title = NULL))
}
```

Mydata contains ALL cross-sectional and longitudinal data across all phases for 1) Cattell, 2) STW, 3) Memory Imm, 4) Memory Del, 5) Simple Spped (inv RT Simple), 6) Choice Speed (inv RT choice). 

```{r}
mydata <- read.csv("CambridgeCentreForAg-AllCognitiveData_DATA_2026-05-17_1254.csv")
```

#Data wrangling 
```{r}
mydata$phase <- ifelse(grepl("phase_1", mydata$redcap_event_name), "1", 
                ifelse(grepl("phase_2", mydata$redcap_event_name), "2", 
                ifelse(grepl("phase_3", mydata$redcap_event_name), "3", 
                ifelse(grepl("phase_4", mydata$redcap_event_name), "4",
                ifelse(grepl("phase_5", mydata$redcap_event_name), "5", 
                ifelse(grepl("participant_info", mydata$redcap_event_name), "999", NA))))))

names(mydata)[names(mydata) == "record_id"] <- "CC.ID"
names(mydata)[names(mydata) == "question_num_years_edu"] <- "education_years"
names(mydata)[names(mydata) == "phase_cog_age"] <- "Age"
names(mydata)[names(mydata) == "cattell_total_score"] <- "Cattell"
names(mydata)[names(mydata) == "stw_n_correct_trials"] <- "STW"
names(mydata)[names(mydata) == "story_memory_immediate_score"] <- "mem_imm"
names(mydata)[names(mydata) == "story_memory_delayed_score"] <- "mem_del"
names(mydata)[names(mydata) == "story_memory_recog_score"] <- "mem_recog"
names(mydata)[names(mydata) == "rt_simple_inv_rt_mean"] <- "simple_speed"
names(mydata)[names(mydata) == "rt_choice_inv_rt_mean"] <- "choice_speed"
names(mydata)[names(mydata) == "phase_cog_online_paper"] <- "format"

mydata$education_years <- as.numeric(mydata$education_years)

mydata <- mydata %>%
  group_by(CC.ID) %>%
  mutate(education_years = {
    val <- education_years[phase == 1 & !is.na(education_years)]
    if (length(val) > 0) val[1] else education_years
  }) %>%
  ungroup()

mydata <- mydata %>%
  group_by(CC.ID) %>%
  mutate(across(c(sex, dob, phase_code), ~ {
    val <- .x[phase == 999 & !is.na(.x)]
    if (length(val) > 0) val[1] else .x
  })) %>%
  ungroup()

mydata <- mydata %>% filter(phase != 999 | is.na(phase))

cog_measures <- c("Cattell", "STW", "mem_imm", "mem_del",
                  "simple_speed", "choice_speed")
mydata <- mydata %>% mutate(across(all_of(cog_measures), as.numeric))

mydata$wave <- NA
mydata$wave[mydata$phase == 1] <- 0
mydata$wave[mydata$phase == 2] <- 0
mydata$wave[mydata$phase == 4] <- 1
mydata$wave[mydata$phase == 5] <- 2

mydata$mem_imm[mydata$CC.ID == 620697 & mydata$wave == 2] <- NA #REDCAP ENTRY MISTAKE 
mydata$mem_del[mydata$CC.ID == 620697 & mydata$wave == 2] <- NA #REDCAP ENTRY MISTAKE 

subject_patterns <- mydata %>%
  mutate(
    participated = as.integer(if_any(all_of(cog_measures), ~ !is.na(.x)))
  ) %>%
  group_by(CC.ID) %>%
  summarize(
    wave0 = as.integer(any(participated == 1 & wave == 0)),
    wave1 = as.integer(any(participated == 1 & wave == 1)),
    wave2 = as.integer(any(participated == 1 & wave == 2)),
    .groups = "drop"
  ) %>%
  mutate(
    pattern = paste0(wave0, wave1, wave2),
    group = case_when(
      pattern == "100"             ~ "G1",
      pattern == "110"             ~ "G2",
      pattern %in% c("111", "101") ~ "G3",
      TRUE                         ~ "Other"
    )
  )

table(subject_patterns$pattern, useNA = "ifany")
table(subject_patterns$group,   useNA = "ifany")
mydata <- merge(mydata, subject_patterns, by = "CC.ID")

missing_ids <- read.csv("missing_record_ids.csv")
names(missing_ids)[names(missing_ids) == "record_id"] <- "CC.ID"
mydata <- mydata[!mydata$CC.ID %in% missing_ids$CC.ID, ]
mydata %>%  group_by(group) %>% summarise(n_unique = n_distinct(CC.ID, na.rm = FALSE))
mydata$return_code <- ifelse(mydata$group == "G3", 1, 0)
mydata$return_code <- as.factor(mydata$return_code)

mydata$dob            <- as.Date(mydata$dob)
mydata$phase_cog_date <- as.Date(mydata$phase_cog_date)
mydata$Age            <- round(as.numeric((mydata$phase_cog_date - mydata$dob) / 365.25), 2)
mydata$sex            <- as.factor(mydata$sex)
```

```{r}
CC700 <- read.csv("CC700_CC.ID.csv")
names(CC700)[names(CC700) == "record_id"] <- "CC.ID"

mydata700 <- mydata[mydata$CC.ID %in% CC700$CC.ID, ]

mydata1_R_NR <- mydata700[mydata700$phase == 1, ]
mydata2_R_NR <- mydata700[mydata700$phase == 2, ]
```

```{r, warning=FALSE}
p1 <- plot_violin(mydata1_R_NR, "STW", "return_code")
p2 <- plot_violin(mydata2_R_NR, "Cattell", "return_code")
p3 <- plot_violin(mydata1_R_NR, "mem_imm", "return_code")
p4 <- plot_violin(mydata1_R_NR, "mem_del", "return_code")
p5 <- plot_violin(mydata1_R_NR, "simple_speed", "return_code")
p6 <- plot_violin(mydata1_R_NR, "choice_speed", "return_code")
grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)

#Keep raw values for variables that will be transformed 
mydata$STW_raw <- mydata$STW
mydata$STW     <- (mydata$STW_raw)^3

#check normality after transformation 
p1 <- plot_violin(mydata1_R_NR, "STW", "return_code")
p2 <- plot_violin(mydata2_R_NR, "Cattell", "return_code")
p3 <- plot_violin(mydata1_R_NR, "mem_imm", "return_code")
p4 <- plot_violin(mydata1_R_NR, "mem_del", "return_code")
p5 <- plot_violin(mydata1_R_NR, "simple_speed", "return_code")
p6 <- plot_violin(mydata1_R_NR, "choice_speed", "return_code")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```

## Plot violins without outliers
```{r}
# p1 <- plot_violin_outlier_suppress(mydata1_R_NR, "STW", "return_code")
# p2 <- plot_violin_outlier_suppress(mydata2_R_NR, "Cattell", "return_code")
# p3 <- plot_violin_outlier_suppress(mydata1_R_NR, "mem_imm", "return_code")
# p4 <- plot_violin_outlier_suppress(mydata1_R_NR, "mem_del", "return_code")
# p5 <- plot_violin_outlier_suppress(mydata1_R_NR, "simple_speed", "return_code")
# p6 <- plot_violin_outlier_suppress(mydata1_R_NR, "choice_speed", "return_code")

# grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```

Phase separation between Phase 1 and Phase 2 is a pratical helper due to Redcap construction. We treat cognitive variables collected in Phase 1 as Phase 2 interchangeably (due to a very short time lag). 

```{r}
mydata1_R <- mydata1_R_NR[mydata1_R_NR$group == "G3", ]
mydata2_R <- mydata2_R_NR[mydata2_R_NR$group == "G3", ]

mydata1_NR <- mydata1_R_NR[mydata1_R_NR$group != "G3" & !is.na(mydata1_R_NR$CC.ID), ]
mydata2_NR <- mydata2_R_NR[mydata2_R_NR$group != "G3" & !is.na(mydata2_R_NR$CC.ID), ]
```

```{r}
g3_in_phase1 <- data.frame(CC.ID = unique(mydata$CC.ID[mydata$group == "G3"]))

CCTina <- read.csv("CC.ID_Tina.csv")
setdiff(CCTina$CC.ID, g3_in_phase1$CC.ID)
setdiff(g3_in_phase1$CC.ID, mydata1_R$CC.ID)
```

# AGE AND SEX DISTRIBUTION 
```{r}
summary(mydata$Age)
table(mydata$sex)         # 2 - female #1 - male

summary(mydata1_R$Age)
summary(mydata1_NR$Age)

table(mydata1_R$sex)
table(mydata1_NR$sex)

sex_ratio <- table(mydata1_R_NR$return_code, mydata1_R_NR$sex)
print(sex_ratio)
chisq.test(sex_ratio)

m_age <- glm(return_code ~ Age, data = mydata1_R_NR, na.action = na.omit, family = binomial)
anova(m_age, test = "Chisq")
m_sex <- glm(return_code ~ sex, data = mydata1_R_NR, na.action = na.omit, family = binomial)
anova(m_sex, test = "Chisq")
m_edu <- glm(return_code ~ education_years, data = mydata1_R_NR, na.action = na.omit, family = binomial)
anova(m_edu, test = "Chisq")
m_full <- glm(return_code ~ sex*Age, data = mydata1_R_NR, na.action = na.omit, family = binomial)
anova(m_full, test = "Chisq")
```

```{r, warning = FALSE}
ggplot(mydata1_R_NR, aes(x = Age, fill = return_code)) +
  geom_histogram(binwidth = 4, position = "dodge", color = "black", size = 0.2) +
  labs(x = "Age", y = "N", fill = "Group") +
  ggtitle("Baseline Age Distribution in NR and R") +
  scale_fill_manual(values = c("lightblue", "darkorange"),
                    labels = c("0" = "NR", "1" = "R"), name = NULL) + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12), 
        legend.position = "bottom") +
  xlim(0, 110)

mydata1_R <- mydata1_R %>%
  mutate(Age = as.numeric(Age),  
         N = ifelse(sex == 1, -1, 1)) 

ggplot(mydata1_R, aes(x = Age, fill = factor(sex))) +
  geom_histogram(data = filter(mydata1_R, sex == 1), aes(y = -..count..), 
                 binwidth = 4, color = "black", size = 0.2, position = "identity") +
  geom_histogram(data = filter(mydata1_R, sex == 2), aes(y = ..count..), 
                 binwidth = 4, color = "black", size = 0.2, position = "identity") +
  scale_fill_manual(values = c("#9BCD9B", "#FFE4E1"),
                    labels = c("1" = "Male", "2" = "Female"), name = NULL) + 
  labs(x = "Age", y = "N", fill = "sex") +
  ggtitle("Age Distribution by Sex in R at a Baseline") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14), 
        legend.position = "bottom") + 
  xlim(0, 110) +
  coord_flip() +  # Flip coordinates for a horizontal layout
  scale_y_continuous(labels = abs, limits = c(-15, 15))  # Combine labels and limits
```

# EDUCATION  
```{r}
outliers_edu    <- boxplot.stats(mydata1_R_NR$education_years)$out
outliers_edu_R  <- boxplot.stats(mydata1_R$education_years)$out
outliers_edu_NR <- boxplot.stats(mydata1_NR$education_years)$out

mydata1_R_NR$education_years[mydata1_R_NR$return_code == 1 & mydata1_R_NR$education_years %in% outliers_edu_R] <- NA
mydata1_R_NR$education_years[mydata1_R_NR$return_code == 0 & mydata1_R_NR$education_years %in% outliers_edu_NR] <- NA

wilcox.test(education_years ~ return_code, data = mydata1_R_NR)

summary(mydata1_R$Education)
summary(mydata1_NR$Education)

ggplot(mydata1_R_NR, aes(x = factor(return_code), y = education_years, fill = factor(return_code))) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Box plot with some transparency
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "red", fill = "white") + 
  labs(title = "Years of Education",
       y = "Years of Education", 
       x = NULL) +
  scale_fill_manual(values = c("0" = "lightblue", "1" = "darkorange"), 
                    labels = c("0" = "NR", "1" = "R"), 
                    name = NULL) +
  scale_x_discrete(labels = c("0" = "NR", "1" = "R")) + 
  theme_minimal() + 
  theme(
      legend.position = "bottom",
      legend.background = element_rect(fill = "transparent", color = NA),
      plot.title = element_text(hjust = 0.5, vjust = 3, size = 12)
    ) + 
  ylim(5, 20)
```

## Education selectivity 
```{r}
clean_data <- na.omit(mydata1_R_NR[, c("education_years", "Age", "return_code")])
education_model <- lm(education_years ~ poly(Age,2), data = clean_data)
clean_data$regressed_education <- residuals(education_model, na.rm = TRUE)

r_residuals_ed <- clean_data$regressed_education[clean_data$return_code == "1"]
cc700_residuals_ed <- clean_data$regressed_education

(mean(r_residuals_ed) - (mean(cc700_residuals_ed)))/sd(cc700_residuals_ed)
(mean(mydata1_R$education_years, na.rm = TRUE) - (mean(mydata1_R_NR$education_years, na.rm = TRUE)))/sd(mydata1_R_NR$education_years, na.rm = TRUE)

clean_data <- na.omit(mydata1_R_NR[, c("education_years", "sex", "return_code", "Age")])
m_education <-lm(education_years ~ return_code * poly(Age, 2), data = clean_data)
summary(m_education)
car::Anova(m_education, type=3)
```

# FLUID INTELLIGENCE (CATTELL) 
```{r}
ggplot(mydata2_R_NR, aes(x = Age, y = Cattell, color = return_code)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = , size = 1) +
  labs(x = "Age", y = "Cattell") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.grid.major = element_line(color = "grey85", size = 0.5),
    panel.grid.minor = element_line(color = "grey90", size = 0.25)
  ) +
  scale_color_manual(values = c("darkorange", "#570861"), labels = c("0" = "Not Returned", "1" = "Returned"), name = NULL
)

outliers_cattell_R <- boxplot.stats(mydata2_R$Cattell)$out
outliers_cattell_NR <- boxplot.stats(mydata2_NR$Cattell)$out

mydata2_R_NR$Cattell[mydata2_R_NR$return_code == 1 & mydata2_R_NR$Cattell %in% outliers_cattell_R] <- NA
mydata2_R_NR$Cattell[mydata2_R_NR$return_code == 0 & mydata2_R_NR$Cattell %in% outliers_cattell_NR] <- NA

plot_violin(mydata2_R_NR, "Cattell", "return_code")

t.test(mydata2_R$Cattell, mydata2_NR$Cattell, paired = FALSE)
wilcox.test(Cattell ~ return_code, data = mydata2_R_NR)
```

## Cattell Selectivity 
```{r}
clean_data <- na.omit(mydata2_R_NR[, c("Cattell", "Age", "education_years", "return_code", "sex")])
clean_data$return_code <- as.factor(clean_data$return_code)
clean_data$sex <- as.factor(clean_data$sex)

cat_model <- lm(Cattell ~ poly(Age,2), data = clean_data)
clean_data$regressed_cattell <- residuals(cat_model, na.rm = TRUE)

returners_residuals <- clean_data$regressed_cattell[clean_data$return_code == "1"]
cc700_residuals <- clean_data$regressed_cattell

#age-orthogonal selectivity statistic
(mean(returners_residuals) - mean(cc700_residuals)) / sd(cc700_residuals)

#age-driven selectivity statistic
(mean(mydata2_R$Cattell, na.rm = TRUE) - (mean(mydata2_R_NR$Cattell, na.rm = TRUE)))/sd(mydata2_R_NR$Cattell, na.rm = TRUE)

m_cattell <-lm(Cattell ~ return_code * scale(education_years) * poly(Age, 2) * sex, data = clean_data)
summary(m_cattell)
car::Anova(m_cattell, type=3)
```

## Cattell Attrition (Pattern-mixture) & ICC
```{r}
df_cattell <- mydata %>% 
  filter(group %in% c("G2", "G3"),
         phase %in% c(2, 4, 5)) %>% 
  group_by(CC.ID) %>% 
  filter(sum(!is.na(Cattell)) >= 2) %>% 
  ungroup()

df_cattell <- df_cattell %>%
  group_by(CC.ID) %>%
  mutate(A0 = as.numeric(Age[wave == '0']), 
         dA = as.numeric(Age - A0))

df_cattell$prac <- 1
df_cattell$prac[str_detect(df_cattell$wave, "0")] <- 0
df_cattell$prac   <- factor(df_cattell$prac)
df_cattell$format <- factor(df_cattell$format)
df_cattell$format <- ifelse(df_cattell$wave == 0 & is.na(df_cattell$format), 1, df_cattell$format)

CC.ID <- df_cattell$CC.ID[df_cattell$wave == 0]
C0    <- df_cattell$Cattell[df_cattell$wave == 0]
C1    <- df_cattell$Cattell[df_cattell$wave == 1]
C2    <- df_cattell$Cattell[df_cattell$wave == 2]
educ  <- df_cattell$education_years[df_cattell$wave == 0]
sex   <- df_cattell$sex[df_cattell$wave == 0]
A0    <- df_cattell$Age[df_cattell$wave == 0]

feature_matrix <- data.frame(CC.ID, C0, C1, C2, A0, sex, educ)

set.seed(987)
iso <- isolation.forest(feature_matrix[, c("C0", "C1", "C2", "A0")], ntrees = 100, sample_size = 128, nthreads = 1)

feature_matrix$pred <- predict(iso, feature_matrix[, c("C0", "C1", "C2", "A0")])
outliers <- feature_matrix[feature_matrix$pred > 0.60, c("CC.ID", "C0", "C1", "C2", "A0", "pred")]
outliers_cattell <- outliers

plot_outliers(df_cattell, "Age", "Cattell", "Age", "Cattell", min(df_cattell$Cattell), max(df_cattell$Cattell), outliers = outliers_cattell)

df_cattell$cattell[df_cattell$CC.ID %in% outliers_cattell$CC.ID] <- NA
df_cattell$cattell_scaled[df_cattell$CC.ID %in% outliers_cattell$CC.ID] <- NA

m_cattell = lmer('scale(Cattell) ~ scale(A0)*scale(dA)*group + scale(A0)*scale(education_years)*sex + format + prac + scale(A0):prac + (1|CC.ID)', data = df_cattell, REML = TRUE)
summary(m_cattell)

wide <- pivot_wider(df_cattell[df_cattell$group == "G3" & df_cattell$wave %in% c(0, 2),],
  id_cols = CC.ID,
  names_from = wave,
  values_from = Cattell)
mat <- wide[, -1]
icc(mat, model = "twoway", type = "consistency", unit = "single")
```

# STW 
```{r}
ggplot(mydata1_R_NR, aes(x = Age, y = STW, color = return_code)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = , size = 1) +
  labs(x = "Age", y = "STW") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.grid.major = element_line(color = "grey85", size = 0.5),
    panel.grid.minor = element_line(color = "grey90", size = 0.25)
  ) +
  scale_color_manual(values = c("darkorange", "#570861"), labels = c("0" = "Not Returned", "1" = "Returned"), 
                     name = NULL
)

sum(is.na(mydata1_R_NR$STW) == TRUE) #N = 2 missing 
sum(is.na(mydata1_R$STW) == TRUE)    #N = 0 missing 
sum(is.na(mydata1_NR$STW) == TRUE)   #N = 2 missing 

outliers_stw_R <- boxplot.stats(mydata1_R$STW)$out
outliers_stw_NR <- boxplot.stats(mydata1_NR$STW)$out

mydata1_R_NR$STW[mydata1_R_NR$return_code == 1 & mydata1_R_NR$STW %in% outliers_stw_R] <- NA
mydata1_R_NR$STW[mydata1_R_NR$return_code == 0 & mydata1_R_NR$STW %in% outliers_stw_NR] <- NA

mydata1_R <- mydata1_R_NR[mydata1_R_NR$phase == 1 & mydata1_R_NR$return_code == 1, ]
mydata1_NR <- mydata1_R_NR[mydata1_R_NR$phase == 1 & mydata1_R_NR$return_code == 0, ]

sum(!is.na(mydata1_R_NR$STW) == TRUE)
sum(!is.na(mydata1_R$STW) == TRUE) 
sum(!is.na(mydata1_NR$STW) == TRUE) 

t.test(mydata1_R$STW, mydata1_NR$STW, paired = FALSE)
wilcox.test(STW ~ return_code, data = mydata1_R_NR)
```

## STW Selectivity 
```{r}
clean_data <- na.omit(mydata1_R_NR[, c("STW", "Age", "education_years", "return_code", "sex")])
clean_data$return_code <- as.factor(clean_data$return_code)
clean_data$sex <- as.factor(clean_data$sex)

stw_model <- lm(STW ~ poly(Age,2), data = clean_data)
clean_data$regressed_stw <- residuals(stw_model, na.rm = TRUE)

returners_residuals <- clean_data$regressed_stw[clean_data$return_code == "1"]
cc700_residuals     <- clean_data$regressed_stw

(mean(returners_residuals) - mean(cc700_residuals)) / sd(cc700_residuals)
(mean(mydata1_R$STW, na.rm = TRUE) - (mean(mydata1_R_NR$STW, na.rm = TRUE)))/sd(mydata1_R_NR$STW, na.rm = TRUE)

m_stw <-lm(STW ~ return_code*poly(Age,2)*scale(education_years)*sex , data = clean_data)
summary(m_stw)
car::Anova(m_stw, type=3)
```
## STW Attrition and ICC
```{r}
df_stw <- mydata %>% 
  filter(group %in% c("G2", "G3"),
         phase %in% c(1, 4, 5)) %>% 
  group_by(CC.ID) %>% 
  filter(sum(!is.na(STW)) >= 2) %>% 
  ungroup()

df_stw <- df_stw %>%
  group_by(CC.ID) %>%
  mutate(A0 = as.numeric(Age[wave == '0']), 
         dA = as.numeric(Age - A0))

df_stw$prac <- 1
df_stw$prac[str_detect(df_stw$wave, "0")] <- 0
df_stw$prac   <- factor(df_stw$prac)
df_stw$format <- factor(df_stw$format)
df_stw$format <- ifelse(df_stw$wave == 0 & is.na(df_stw$format), 1, df_stw$format)

CC.ID <- df_stw$CC.ID[df_stw$wave == 0]
C0    <- df_stw$STW[df_stw$wave == 0]
C1    <- df_stw$STW[df_stw$wave == 1]
C2    <- df_stw$STW[df_stw$wave == 2]
educ  <- df_stw$education_years[df_stw$wave == 0]
sex   <- df_stw$sex[df_stw$wave == 0]
A0    <- df_stw$Age[df_stw$wave == 0]

feature_matrix <- data.frame(CC.ID, C0, C1, C2, A0, sex, educ)

set.seed(987)
iso <- isolation.forest(feature_matrix[, c("C0", "C1", "C2", "A0")], ntrees = 100, sample_size = 128, nthreads = 1)

feature_matrix$pred <- predict(iso, feature_matrix[, c("C0", "C1", "C2", "A0")])
outliers <- feature_matrix[feature_matrix$pred > 0.60, c("CC.ID", "C0", "C1", "C2", "A0", "pred")]
outliers_stw <- outliers

plot_outliers(df_stw, "Age", "STW", "Age", "STW", min(df_stw$STW), max(df_stw$STW), outliers = outliers_stw)

df_stw$STW[df_stw$CC.ID %in% outliers_stw$CC.ID] <- NA

m_STW = lmer('scale(STW) ~ scale(A0)*scale(dA)*group + scale(A0)*scale(education_years)*sex + format + prac + scale(A0):prac + (1|CC.ID)', data = df_stw, REML = TRUE)
summary(m_STW)

wide <- pivot_wider(df_stw[df_stw$group == "G3" & df_stw$wave %in% c(0, 2),],
  id_cols = CC.ID,
  names_from = wave,
  values_from = STW)
mat <- wide[, -1]
icc(mat, model = "twoway", type = "consistency", unit = "single")

#sensitivity check
df_sub <- df_stw[df_stw$group == "G3", ]
mod <- lmer(STW ~ 1 + (1 | CC.ID), data = df_sub)
vc <- as.data.frame(VarCorr(mod))
ICC_C <- vc$vcov[vc$grp == "CC.ID"] /
         (vc$vcov[vc$grp == "CC.ID"] + vc$vcov[vc$grp == "Residual"])
ICC_C
```

# VERBAL EPISODIC MEMORY 
```{r}
ggplot(mydata1_R_NR, aes(x = Age, y = mem_imm, color = return_code)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE , size = 1) +
  labs(x = "Age", y = "Memory Adj") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.grid.major = element_line(color = "grey85", size = 0.5),
    panel.grid.minor = element_line(color = "grey90", size = 0.25)
  ) +
  scale_color_manual(values = c("darkorange", "#570861"), labels = c("0" = "Not Returned", "1" = "Returned"),
                     name = NULL
)

ggplot(mydata1_R_NR, aes(x = Age, y = mem_del, color = return_code)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE , size = 1) +
  labs(x = "Age", y = "Memory Adj") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.grid.major = element_line(color = "grey85", size = 0.5),
    panel.grid.minor = element_line(color = "grey90", size = 0.25)
  ) +
  scale_color_manual(values = c("darkorange", "#570861"), labels = c("0" = "Not Returned", "1" = "Returned"),
                     name = NULL
)

outliers_mem_imm_R  <- boxplot.stats(mydata1_R$mem_imm)$out
outliers_mem_imm_NR <- boxplot.stats(mydata1_NR$mem_imm)$out

mydata1_R_NR$mem_imm[mydata1_R_NR$return_code == 1 & mydata1_R_NR$mem_imm %in% outliers_mem_imm_R] <- NA
mydata1_R_NR$mem_imm[mydata1_R_NR$return_code == 0 & mydata1_R_NR$mem_imm %in% outliers_mem_imm_NR] <- NA

outliers_mem_del_R  <- boxplot.stats(mydata1_R$mem_del)$out
outliers_mem_del_NR <- boxplot.stats(mydata1_NR$mem_del)$out

mydata1_R_NR$mem_del[mydata1_R_NR$return_code == 1 & mydata1_R_NR$mem_del %in% outliers_mem_del_R] <- NA
mydata1_R_NR$mem_del[mydata1_R_NR$return_code == 0 & mydata1_R_NR$mem_del %in% outliers_mem_del_NR] <- NA

wilcox.test(mem_imm ~ return_code, data = mydata1_R_NR)
wilcox.test(mem_del ~ return_code, data = mydata1_R_NR)
```

## Memory Selectivity
```{r}
#selectivity age-driven 
(mean(mydata1_R$mem_imm, na.rm = TRUE) - (mean(mydata1_R_NR$mem_imm, na.rm = TRUE)))/sd(mydata1_R_NR$mem_imm, na.rm = TRUE)
(mean(mydata1_R$mem_del, na.rm = TRUE) - (mean(mydata1_R_NR$mem_del, na.rm = TRUE)))/sd(mydata1_R_NR$mem_del, na.rm = TRUE)

# selectivity age-orthogonal (and education-adjusted) for Imm
clean_data <- na.omit(mydata1_R_NR[, c("mem_imm", "Age", "education_years", "return_code")])
imm_model <- lm(mem_imm ~ poly(Age,2), data = clean_data)
clean_data$regressed_imm <- residuals(imm_model, na.rm = TRUE)

returners_residuals <- clean_data$regressed_imm[clean_data$return_code == "1"]
cc700_residuals     <- clean_data$regressed_imm
(mean(returners_residuals) - mean(cc700_residuals)) / sd(cc700_residuals)

m_mem <-lm(mem_imm ~ return_code*poly(Age,2)*scale(education_years) , data = clean_data)
summary(m_mem)
car::Anova(m_mem, type=3, statistic = F)

# selectivity age-orthogonal (and education-adjusted) for Del
clean_data <- na.omit(mydata1_R_NR[, c("mem_del", "Age", "education_years", "return_code")])
del_model <- lm(mem_del ~ poly(Age,2), data = clean_data)
clean_data$regressed_del <- residuals(del_model, na.rm = TRUE)

returners_residuals <- clean_data$regressed_del[clean_data$return_code == "1"]
cc700_residuals     <- clean_data$regressed_del
(mean(returners_residuals) - mean(cc700_residuals)) / sd(cc700_residuals)

m_mem <-lm(mem_del ~ return_code*poly(Age,2)*scale(education_years) , data = clean_data)
summary(m_mem)
car::Anova(m_mem, type=3, statistic = F)
```
## Memory ICC 
```{r}
df_mem <- mydata %>% 
  filter(group %in% c("G2", "G3"),
         phase %in% c(1, 4, 5)) %>% 
  group_by(CC.ID) %>% 
  filter(sum(!is.na(mem_imm) & !is.na(mem_del)) >= 2) %>% 
  ungroup()

df_mem <- df_mem %>%
  group_by(CC.ID) %>%
  mutate(A0 = as.numeric(Age[wave == '0']), 
         dA = as.numeric(Age - A0))

df_mem$prac <- 1
df_mem$prac[str_detect(df_mem$wave, "0")] <- 0
df_mem$prac   <- factor(df_mem$prac)
df_mem$format <- factor(df_mem$format)
df_mem$format <- ifelse(df_mem$wave == 0 & is.na(df_mem$format), 1, df_mem$format)

CC.ID <- df_mem$CC.ID[df_mem$wave == 0]
C0    <- df_mem$mem_imm[df_mem$wave == 0]
C1    <- df_mem$mem_imm[df_mem$wave == 1]
C2    <- df_mem$mem_imm[df_mem$wave == 2]
C0_1  <- df_mem$mem_del[df_mem$wave == 0]
C1_1  <- df_mem$mem_del[df_mem$wave == 1]
C2_1  <- df_mem$mem_del[df_mem$wave == 2]
educ  <- df_mem$education_years[df_mem$wave == 0]
sex   <- df_mem$sex[df_mem$wave == 0]
A0    <- df_mem$Age[df_mem$wave == 0]

feature_matrix <- data.frame(CC.ID, C0, C1, C2, C0_1, C1_1, C2_1, A0, sex, educ)

set.seed(987)
iso <- isolation.forest(feature_matrix[, c("C0", "C1", "C2", "C0_1", "C1_1", "C2_1", "A0")], ntrees = 100, sample_size = 128, nthreads = 1)

feature_matrix$pred <- predict(iso, feature_matrix[, c("C0", "C1", "C2", "C0_1", "C1_1", "C2_1", "A0")])
outliers <- feature_matrix[feature_matrix$pred > 0.60, c("CC.ID", "C0", "C1", "C2", "C0_1", "C1_1", "C2_1", "A0", "pred")]
outliers_mem <- outliers

plot_outliers(df_mem, "Age", "mem_imm", "Age", "Memory Imm", min(df_mem$mem_imm), max(df_mem$mem_imm), outliers = outliers_mem)
plot_outliers(df_mem, "Age", "mem_imm", "Age", "Memory Del", min(df_mem$mem_del), max(df_mem$mem_del), outliers = outliers_mem)

df_mem$mem_imm[df_mem$CC.ID %in% outliers_mem$CC.ID] <- NA
df_mem$mem_del[df_mem$CC.ID %in% outliers_mem$CC.ID] <- NA

m_imm = lmer('scale(mem_imm) ~ scale(A0)*scale(dA)*group + scale(A0)*scale(education_years)*sex + format + prac + scale(A0):prac + (1|CC.ID)', data = df_mem, REML = TRUE)
summary(m_imm)

wide <- pivot_wider(df_mem[df_mem$group == "G3" & df_mem$wave %in% c(0, 2),],
  id_cols = CC.ID,
  names_from = wave,
  values_from = mem_imm)
mat <- wide[, -1]
icc(mat, model = "twoway", type = "consistency", unit = "single")

wide <- pivot_wider(df_mem[df_mem$group == "G3" & df_mem$wave %in% c(0, 2),],
  id_cols = CC.ID,
  names_from = wave,
  values_from = mem_del)
mat <- wide[, -1]
icc(mat, model = "twoway", type = "consistency", unit = "single")

# #sensitivity check
# df_sub <- df_mem[df_mem$group == "G3", ]
# mod <- lmer(mem_imm ~ 1 + (1 | CC.ID), data = df_sub)
# summary(mod)
# vc <- as.data.frame(VarCorr(mod))
# ICC_C <- vc$vcov[vc$grp == "CC.ID"] /
#          (vc$vcov[vc$grp == "CC.ID"] + vc$vcov[vc$grp == "Residual"])
# ICC_C
# 
# #sensitivity check
# mod <- lmer(mem_imm ~ 1 + (1 | CC.ID), data = df_mem)
# summary(mod)
# vc <- as.data.frame(VarCorr(mod))
# var_between <- vc$vcov[vc$grp == "CC.ID"]
# var_within  <- vc$vcov[vc$grp == "Residual"]
# ICC_C <- var_between / (var_between + var_within)
# ICC_C
```

# SPEED (inverse Reaction Time)
```{r}
# outliers RT
outliers_rt_simple_R <- boxplot.stats(mydata1_R$simple_speed)$out
outliers_rt_simple_NR <- boxplot.stats(mydata1_NR$simple_speed)$out

mydata1_R_NR$simple_speed[mydata1_R_NR$return_code == 1 & mydata1_R_NR$simple_speed %in% outliers_rt_simple_R] <- NA
mydata1_R_NR$simple_speed[mydata1_R_NR$return_code == 0 & mydata1_R_NR$simple_speed %in% outliers_rt_simple_NR] <- NA

outliers_rt_choice_R <- boxplot.stats(mydata1_R$choice_speed)$out
outliers_rt_choice_NR <- boxplot.stats(mydata1_NR$choice_speed)$out

mydata1_R_NR$choice_speed[mydata1_R_NR$return_code == 1 & mydata1_R_NR$choice_speed %in% outliers_rt_choice_R] <- NA
mydata1_R_NR$choice_speed[mydata1_R_NR$return_code == 0 & mydata1_R_NR$choice_speed %in% outliers_rt_choice_NR] <- NA

wilcox.test(simple_speed ~ return_code, data = mydata1_R_NR)
wilcox.test(choice_speed ~ return_code, data = mydata1_R_NR)
```

## Speed Selectivity
```{r}
# selectivity age-driven
(mean(mydata1_R$simple_speed, na.rm = TRUE) - (mean(mydata1_R_NR$simple_speed, na.rm = TRUE)))/sd(mydata1_R_NR$simple_speed, na.rm = TRUE)
(mean(mydata1_R$choice_speed, na.rm = TRUE) - (mean(mydata1_R_NR$choice_speed, na.rm = TRUE)))/sd(mydata1_R_NR$choice_speed, na.rm = TRUE)

# selectivity age-orthogonal
clean_data <- na.omit(mydata1_R_NR[, c("simple_speed", "education_years", "return_code", "Age", "sex")])

rt_model <- lm(simple_speed ~ poly(Age,2), data = clean_data)
clean_data$regressed_simple <- residuals(rt_model, na.rm = TRUE)
returners_residuals <- clean_data$regressed_simple[clean_data$return_code == "1"]
cc700_residuals <- clean_data$regressed_simple
(mean(returners_residuals) - mean(cc700_residuals)) / sd(cc700_residuals)

rt_m <-lm(simple_speed ~ return_code*poly(Age,2)*scale(education_years) , data = clean_data)
summary(rt_m)
car::Anova(rt_m, type=3, statistic = F)

# selectivity age-orthogonal choice
clean_data <- na.omit(mydata1_R_NR[, c("choice_speed", "education_years", "return_code", "Age")])

rt_model <- lm(choice_speed ~ poly(Age,2), data = clean_data)
clean_data$regressed_choice <- residuals(rt_model, na.rm = TRUE)
returners_residuals <- clean_data$regressed_choice[clean_data$return_code == "1"]
cc700_residuals <- clean_data$regressed_choice
(mean(returners_residuals) - mean(cc700_residuals)) / sd(cc700_residuals)

rt_m <-lm(choice_speed ~ return_code*poly(Age,2)*scale(education_years) , data = clean_data)
summary(rt_m)
car::Anova(rt_m, type=3, statistic = F)
```

## Speed ICC
```{r}
df_speed <- mydata %>% 
  filter(group %in% c("G2", "G3"),
         phase %in% c(1, 4, 5)) %>% 
  group_by(CC.ID) %>% 
  filter(sum(!is.na(simple_speed) & !is.na(choice_speed)) >= 2) %>% 
  ungroup()

df_speed <- df_speed %>%
  group_by(CC.ID) %>%
  mutate(A0 = as.numeric(Age[wave == '0']), 
         dA = as.numeric(Age - A0))

df_speed$prac <- 1
df_speed$prac[str_detect(df_speed$wave, "0")] <- 0
df_speed$prac   <- factor(df_speed$prac)
df_speed$format <- factor(df_speed$format)
df_speed$format <- ifelse(df_speed$wave == 0 & is.na(df_speed$format), 1, df_speed$format)

CC.ID <- df_speed$CC.ID[df_speed$wave == 0]
C0    <- df_speed$simple_speed[df_speed$wave == 0]
C1    <- df_speed$simple_speed[df_speed$wave == 1]
C2    <- df_speed$simple_speed[df_speed$wave == 2]
C0_1  <- df_speed$choice_speed[df_speed$wave == 0]
C1_1  <- df_speed$choice_speed[df_speed$wave == 1]
C2_1  <- df_speed$choice_speed[df_speed$wave == 2]
educ  <- df_speed$education_years[df_speed$wave == 0]
sex   <- df_speed$sex[df_speed$wave == 0]
A0    <- df_speed$Age[df_speed$wave == 0]

feature_matrix <- data.frame(CC.ID, C0, C1, C2, C0_1, C1_1, C2_1, A0, sex, educ)

set.seed(987)
iso <- isolation.forest(feature_matrix[, c("C0", "C1", "C2", "C0_1", "C1_1", "C2_1", "A0")], ntrees = 100, sample_size = 128, nthreads = 1)

feature_matrix$pred <- predict(iso, feature_matrix[, c("C0", "C1", "C2", "C0_1", "C1_1", "C2_1", "A0")])
outliers <- feature_matrix[feature_matrix$pred > 0.60, c("CC.ID", "C0", "C1", "C2", "C0_1", "C1_1", "C2_1", "A0", "pred")]
outliers_speed <- outliers

plot_outliers(df_speed, "Age", "simple_speed", "Age", "Simple Speed", min(df_speed$simple_speed), max(df_speed$simple_speed), outliers = outliers_speed)
plot_outliers(df_speed, "Age", "choice_speed", "Age", "Choice Speed", min(df_speed$choice_speed), max(df_speed$choice_speed), outliers = outliers_speed)

df_speed$simple_speed[df_speed$CC.ID %in% outliers_speed$CC.ID] <- NA
df_speed$choice_speed[df_speed$CC.ID %in% outliers_speed$CC.ID] <- NA

m_simple = lmer('scale(simple_speed) ~ scale(A0)*scale(dA)*group + scale(A0)*scale(education_years)*sex + format + prac + scale(A0):prac + (1|CC.ID)', data = df_speed, REML = TRUE)
summary(m_simple)

m_choice = lmer('scale(choice_speed) ~ scale(A0)*scale(dA)*group + scale(A0)*scale(education_years)*sex + format + prac + scale(A0):prac + (1|CC.ID)', data = df_speed, REML = TRUE)
summary(m_choice)

wide <- pivot_wider(df_speed[df_speed$group == "G3" & df_speed$wave %in% c(0, 2),],
  id_cols = CC.ID,
  names_from = wave,
  values_from = simple_speed)
mat <- wide[, -1]
icc(mat, model = "twoway", type = "consistency", unit = "single")

wide <- pivot_wider(df_speed[df_speed$group == "G3" & df_speed$wave %in% c(0, 2),],
  id_cols = CC.ID,
  names_from = wave,
  values_from = choice_speed)
mat <- wide[, -1]
icc(mat, model = "twoway", type = "consistency", unit = "single")

```


